---
title: "Independent Study to prepare for workshop"
subtitle: "Omics 3: Visualising and Interpreting"
author: "Emma Rand"
format: 
  revealjs:
    footer: "🔗 [About Omics 3: Visualising and Interpreting](https://3mmarand.github.io/BIO00088H-data/omics/week-5/overview.html)"
    slide-number: true
    chalkboard: true
    code-link: true
    code-line-numbers: false
bibliography: ../../references.bib
editor: 
  markdown: 
    wrap: 72
---

## Overview

In these slides we will:

::: incremental
-   Check where you are

-   learn some concepts 

    -   

-   Find out what packages to install before the workshop
:::

# Where should you be?

## What we did in Omics 2: Statistical Analysis

::: incremental
::: {style="font-size: 90%;"}
-   
-   
-   Saved files .
:::
:::

## Where should you be?

After the [Omics 2: 👋 Statistical Analysis Workshop](../week-4/workshop.html)
including:

-   [🤗 Look after future
    you!](../week-4/workshop.html#look-after-future-you) and

-   the [Independent Study to
    consolidate](../week-4/study_after_workshop.html), you should have:

## 🐸 Frogs

::: {style="font-size: 90%;"}
-   An RStudio Project called `frogs-88H` which contains:
    -   Raw data (S14, S20 and S30)
    -   Processed data (`s30_filtered.csv`, `s30_summary_gene.csv`,
        `s30_summary_gene_filtered.csv`, `s30_summary_samp.csv` and
        equivalents for S14 *OR* S20)
    -   Two scripts called `cont-fgf-s30.R` and `cont-fgf-s20.R` *OR*
        `cont-fgf-s14.R`
:::

Files should be organised into folders. Code should well commented and
easy to read.

## 🐭 Mice

-   An RStudio Project called `mice-88H` which contains
    -   Raw data (hspc, prog, lthsc)
    -   Processed data (`hspc_summary_gene.csv`,
        `hspc_summary_samp.csv`, `prog_summary_gene.csv`,
        `prog_summary_samp.csv`)
-   One script called `hspc-prog.R`

Files should be organised into folders. Code should well commented and
easy to read.

## 🍂

Either of the other examples.

## If you do not have those

Go through:

-   [Omics 2: Statistical Analysis](../week-4/workshop.html)
    including:

-   [🤗 Look after future
    you!](../week-4/workshop.html#look-after-future-you) and

-   the [Independent Study to
    consolidate](../week-4/study_after_workshop.html)

# xx

## xx

::: incremental
-   T
:::



-   The 🐭 mouse data have been normalised to simplify the analysis for
    you; the 🐸 frog data have not but the DE method will do this for
    you.

-
## Packages to install before the workshop

**``** from CRAN in the the normal way:

```{r}
#| eval: false
install.packages("")
```

**`DESeq2`** from Bioconductor using BiocManager:
```{r}
#| eval: false
BiocManager::install("DESeq2")
```



# Workshops

## Workshops

-   Omics 1: Hello data Getting to know the data. Checking the
    distributions of values

-   Omics 2: Statistical Analysis Identifying which genes are
    differentially expressed between treatments.

-   Omics 3: Visualising and Interpreting. PCA, Volcano plots and heatmaps to visualise results. Interpreting the results and finding out more about genes of interest.

## References
