---
title: "Self study"
subtitle: "Semester 2: Additional 🐭 stem cell labels"
author: "Emma Rand"
toc: true
toc-depth: 4
toc-location: right
execute:
  echo: true
  include: true
  error: true
bibliography: ../../references.bib
editor: 
  markdown: 
    wrap: 72
---

# Introduction

We have analysed the @nestorowa2016 stem cells categorised as LTHSC,
HSPC, or Prog. The cell types were determined from flowcytometry
markers. Additional markers allow the cells to be categorised into 12
other types (including unknown): `CMP`, `GMP`, `LMPP`, `LTHSC`,
`LTHSC-ESLAM`, `MEP`, `MPP_broad`, `MPP1`, `MPP2`, `MPP3`, `STHSC`,
`unknown`. These cell types are mutually exclusive, meaning that a cell
can only be one of these types.

In this self-study, you will:

-   import all the expression data and gene information for the three
    cell types
-   import the data that gives the additional labels
-   combine the data and pivot it to have the expression values in a
    single column
-   plot the expression of one gene, `Emb`, in each cell type for the
    3-type categorisation and the 12-type categorisation
-   conduct a PCA on all the data and plot the first two components
    using colour/shapes for the 12-type categorisation, the 3-type
    categorisation and the expression of `Emb`.

# Set up

🎬 Open your `mice-88H` RStudio Project.

🎬 Save [er_cell_types.csv](data-raw/er_cell_types.csv).

🎬 Open a new or existing script

🎬 Load `tidyverse` [@tidyverse].

```{r}
#| echo: false
library(tidyverse)
```

# Import and combine data

## The raw data

Import the data for the LT-HSC, HSPC and the Progenitor cells.

🎬 Import [surfaceome_lthsc.csv](data-raw/surfaceome_lthsc.csv),
[surfaceome_hspc.csv](data-raw/surfaceome_hspc.csv) and
[surfaceome_hspc.csv](data-raw/surfaceome_prog.csv)

```{r}
# 🐭 import the three datasets
lthsc <- read_csv("data-raw/surfaceome_lthsc.csv")
hspc <- read_csv("data-raw/surfaceome_hspc.csv")
prog <- read_csv("data-raw/surfaceome_prog.csv")
```

```{r}
#| echo: false
# number of columns in each dataset
nlthsc <- ncol(lthsc) - 1
nhspc <- ncol(hspc) - 1
nprog <- ncol(prog) - 1
ntotal <- nlthsc + nhspc + nprog
```

Note there are `r nlthsc` LT-HSC + `r nhspc` HSPC + `r nprog` Progenitor
= `r ntotal` total cells.

We will need to join the three datasets using `ensembl_gene_id` to match
the rows.

🎬 Combine the three datasets by `ensembl_gene_id` and save the result
as `cell_expr`.

```{r}
#  combine the three datasets
cell_expr <- hspc |>
  left_join(prog, 
            by = "ensembl_gene_id") |> 
  left_join(lthsc, 
            by = "ensembl_gene_id")
```

## Gene information

You will likely want the gene information in the dataframe along with
the `ensembl_gene_id`. Rather the connecting again to
[Ensembl](https://www.ensembl.org/index.html) [@martin2023; @birney2004]
and [BioMart](https://www.ensembl.org/info/data/biomart/index.html)
[@smedley2009] using the R package **`biomaRt`** [@biomaRt1; @biomaRt2],
which means loading more packages, managing conflicts and using more
compute, we can use the information we already have in the
[results/hspc_prog_results.csv](results/hspc_prog_results.csv) file.
This works because the same 280 genes are in the three raw data files
and this results file.

We need only the `ensembl_gene_id`, (to join the information), the
`external_gene_name` and the `description`.

🎬 Import the gene information from `results/hspc_prog_results.csv` and
save as `gene_info`.

```{r}
# 🐭 import the gene information
gene_info <- read_csv("results/hspc_prog_results.csv") |> 
  select(ensembl_gene_id, external_gene_name, description)
```

🎬 Join the `gene_info` to the `cell_expr` data.

```{r}
# 🐭 join the gene information to the cell_expr data
cell_expr <- cell_expr |> 
  left_join(gene_info, 
            by = "ensembl_gene_id")
```

## 12-type categorisation data

The information in the [er_cell_types.csv](data-raw/er_cell_types.csv)
has 14 columns. The first column gives the cell_id (e.g., `HSPC_001`)
the next 12 columns are named with the cell type or unknown and contain
a 0 or 1. These columns are useful if you want to compare (plot or DE),
for example, CMP cells vs others. The last column gives the cell type 
which is useful to include *all* the labels in a plot.

🎬 Import [er_cell_types.csv](data-raw/er_cell_types.csv)

```{r}
cell <- read_csv("data-raw/er_cell_types.csv")

```

🎬 Check the dataframe has the number of rows and columns you were
expecting and that column types and names are as expected.

🎬 Examine the number of cells of each type:

```{r}
cell |> 
  group_by(cell_type_12) |>
  count()
```

# Expression of a specific gene in each cell type

## Pivot to long form

The `cell_expr` data has the expression of each gene in each cell. The
`cell_ids` are in columns and the gene are in rows.

We will need to pivot the `cell_expr` data to have the expression values
in a single column with additional columns giving the cell_id and gene.

```{r}
# 🐭 pivot the cell_expr data
cell_expr_long <- cell_expr |> 
  pivot_longer(cols = -c(ensembl_gene_id, 
                         external_gene_name, 
                         description),
               names_to = "cell_id", 
               values_to = "expression")
```

Note that this data frame will have:
[280 genes * 1654 cells =  463120] rows.

## Add the two types of categorisation to the expression data

We also want to add the two types of categorisation (the 3-type and the
12-types) to the `cell_expr_long` data. To add the 3-type categorisation
we extract the information in the `cell_id` column into two columns: one
with the cell type and one with the number. We did this before in making
a [PCA plot](../week-5/workshop.html#stem-cells-2)

🎬 Extract the cell type and cell number from the `cell_id` column
(keeping the `cell_id` column):

```{r}
cell_expr_long <- cell_expr_long |> 
  extract(cell_id, 
          remove = FALSE,
          c("cell_type_3", "cell_number"),
          "([a-zA-Z.]{4, 6})_([0-9]{3})")
```

Let's just check we have the right number of cells in each of the three
groups:

```{r}
cell_expr_long |> 
  group_by(cell_type_3) |> 
  summarise(n()/ 280)
```



To add the 12-type categorisation we need to join the `cell_expr_long`
data to the `cell` data on the `cell_id`

🎬 Add the 12-type categorisation to the `cell_expr_long` data.

```{r}
cell_expr_long <- cell_expr_long |> 
  left_join(cell |> select(cell_id, cell_type_12), 
            by = "cell_id")
```

### Plot gene expression under 3-type categorisation

I have chosen the gene `Emb` to plot the expression in each cell type. I
prefer the use of violin plots and points because multimodal
distributions are more obvious than in boxplots

🎬 Plot the expression of `Emb` in each cell type for the 3-type
categorisation.

```{r}
# 🐭 plot the expression of Emb in each cell type
cell_expr_long |> 
  filter(external_gene_name == "Emb") |> 
  ggplot(aes(x = cell_type_3, y = expression)) +
  geom_violin(linewidth = 1) +
  geom_jitter(width = 0.4, alpha = 0.2, pch = 16) +
  theme_classic()
```

### Plot expression under 12-type categorisation

```{r}
# 🐭 plot the expression of Emb in each cell type
cell_expr_long |> 
  filter(external_gene_name == "Emb") |> 
  ggplot(aes(x = cell_type_12, y = expression)) +
  geom_violin(linewidth = 1) +
  geom_jitter(width = 0.2, alpha = 0.2, pch = 16) +
  theme_classic()
```

# PCA

We will carry out the the same steps to do the PCA:

-   Select the log~2~ transformed normalised counts. You can only use
    numerical data in PCA.
-   Transpose our data. We have genes in rows and samples in columns
    (this is common for gene expression data). However, to treat the
    genes as variables, PCA expects samples in rows and genes in
    columns.
-   Add the gene names as column names in the transposed data
-   Perform the PCA
-   Extract the scores on the first two principal components and label
    the data
-   Plot the the first two principal components as a scatter plot

Refer to the [workshop for more explanation on
PCA](../week-5/workshop.html#stem-cells-2).

🎬 Transpose the log~2~ transformed normalised counts:

```{r}
cell_expr_trans <- cell_expr |> 
  dplyr::select(starts_with(c("HSPC_", "Prog_", "LT.HSC_"))) |>
  t() |> 
  data.frame()
```

🎬 Set the column names to the gene ids:

```{r}
colnames(cell_expr_trans) <- cell_expr$ensembl_gene_id
```

🎬 Perform PCA on the log~2~ transformed normalised counts:

```{r}
pca <- cell_expr_trans |>
  prcomp(rank. = 8)
```

To plot the PC1 against PC2 we will need to extract the PC1 and PC2
"scores" from the PCA object and add labels for the cells. Our PCA
object is called `pca` and the scores are in pca\$x. The cells labels
will come from the row names of the transformed data.

🎬 Create a dataframe of the PC1 and PC2 scores (in `pca$x`) and add the
cell ids:

```{r}
pca_labelled <- data.frame(pca$x,
                           cell_id = row.names(cell_expr_trans))
```

🎬 Extract the cell type and cell number from the `cell_id` column
(keeping the `cell_id` column):

```{r}
pca_labelled <- pca_labelled |> 
  extract(cell_id, 
          remove = FALSE,
          c("cell_type_3", "cell_number"),
          "([a-zA-Z.]{4, 6})_([0-9]{3})")
```

🎬 Add the 12-type categorisation to the `pca_labelled` data.

```{r}
pca_labelled <- pca_labelled |> 
  left_join(cell |> select(cell_id, cell_type_12), 
            by = "cell_id")
```

🎬 Add the expression of `Emb` to `pca_labelled` data. I've also renamed
the expression column from the filtered `cell_expr_long` so it is
`Emb_expr` in the `pca_labelled` data.

```{r}
pca_labelled <- pca_labelled |> 
  left_join(cell_expr_long |> 
              filter(external_gene_name == "Emb") |> 
              select(cell_id, "Emb_expr" = expression), 
            by = "cell_id")
```

🎬 Plot PC2 against PC1 and colour by cell_type_3 type:

```{r}
pca_labelled |> 
  ggplot(aes(x = PC1, y = PC2, 
             colour = cell_type_3)) +
  geom_point(alpha = 0.4) +
  scale_colour_viridis_d(end = 0.95, begin = 0.15,
                         name = NULL) +
  theme_classic()

```

🎬 Plot PC2 against PC1 and colour by cell_type_12 type:

```{r}
pca_labelled |> 
  ggplot(aes(x = PC1, y = PC2, 
             colour = cell_type_12)) +
  geom_point(alpha = 0.5) +
  scale_colour_viridis_d(name = NULL) +
  theme_classic()

```

🎬 Plot PC2 against PC1 and colour by expression of Emb and shape by
cell_type_3 type:

```{r}
pca_labelled |> 
  ggplot(aes(x = PC1, y = PC2, 
             colour = Emb_expr,
             shape = cell_type_3)) +
  geom_point() +
  scale_colour_viridis_c(end = 0.95, begin = 0.15,
                         name = NULL) +
  theme_classic()

```

🎬 Easier to see as facets?:

```{r}
pca_labelled |> 
  ggplot(aes(x = PC1, y = PC2, 
             colour = Emb_expr)) +
  geom_point() +
  scale_colour_viridis_c(end = 0.95, begin = 0.15,
                         name = NULL) +
  facet_wrap(~cell_type_3)

```

Pages made with R [@R-core], Quarto [@Allaire_Quarto_2024], `knitr`
[@knitr1; @knitr2; @knitr3], `kableExtra` [@kableExtra]

# References
