---
title: "Workshop"
subtitle: "Omics 1: Hello data!"
toc: true
toc-location: right
---

# Introduction


## Session overview

In this workshop you will 


::: callout-note
## Key


:::

# Getting started

```{r}
library(tidyverse)
```
```
tidyverse_conflicts() ──
✖ dplyr::filter() masks stats::filter()
✖ dplyr::lag()    masks stats::lag()
ℹ Use the conflicted package to force all conflicts to become errors
``` 

See later workshops.

# Exercises

## Set up a Project

-   set up a project and create the organisation


## Examine your data

### Meta data

Frog development: [meta_data.txt](../data/betsy/meta_data.txt)

```{r}
#| echo: false
meta <- read_table("../data/betsy/meta_data.txt")
```

```{r}
#| echo: false
meta |> knitr::kable()

```





### Data
-   open in understanding the data numbers of variables, organisation

## Import

## Explore

-   what kind of values are they, how many missing
-   qc plots
-   removing the useless, writing to files


🎬 

You're finished!

# 🥳 Well Done! 🎉



# Independent study following the workshop

[Consolidate](study_after_workshop.qmd)

# The Code file

These contain all the code needed in the workshop even where it is not visible on the webpage.

The [workshop.qmd](workshop.qmd) file is the file I use to compile the practical. Qmd stands for Quarto markdown. It allows code and ordinary text to be interleaved to produce well-formatted reports including webpages. Right-click on the link and choose Save-As to download. You will be able to open the Qmd file in RStudio. Alternatively, [View in Browser](https://github.com/3mmaRand/). Coding and thinking answers are marked with `#---CODING ANSWER---` and `#---THINKING ANSWER---`

Pages made with R [@R-core], Quarto [@allaire2022], `knitr` [@knitr], `kableExtra` [@kableExtra]

# References
